---
title: "Introduction to R"
subtitle: "For the R-Curious"
format:
  revealjs:
    theme: [default, styles.scss]
    self-contained: true
    chalkboard: false
    standalone: true
    slide-number: true
    show-slide-number: all
knitr:
  opts_chunk:
    dev: png
    dev.args: { bg: "transparent" }
---

```{r}
#| messages: false
#| warning: false

library(tidyverse)
set.seed(3272)
```

## If you don't have R and RStudio

- [r-project.org](https://www.r-project.org/)
    - Download R version for your operating system
- [Posit RStudio](https://posit.co/download/rstudio-desktop)
    - Download RStudio version for your operating system
- Run the installers
    - Defaults okay
- We will use these later.

<hr />

No laptop?

- Sit next to someone -- group work.


## Who are we?

:::: {.columns}

::: {.column width="50%"}
**Dr. Elizabeth King** (she/her)

- Grew up in Southeastern Wisconsin (until 7th grade) $\rightarrow$ Chicago-land
- Iowa (4 years) $\rightarrow$ California (10 years) $\rightarrow$ Missouri
- MU 12 years in Biology
:::

::: {.column width="50%"}
**Dr. Kevin Middleton** (he/him)

- Grew up in Florida
- North Carolina $\rightarrow$ Rhode Island (10 years) $\rightarrow$ California (1 year) $\rightarrow$ Rhode Island (1 year) $\rightarrow$ California (5 years) $\rightarrow$ Missouri
- MU 14 years (10 in Medicine, 4 in Biology)

:::

::::

<hr />

. . .

- Biologists by training
- Engineers, theoreticians, statisticians, and programmers by avocation
- We are married.


## Why are we here?

- Programming/data skills are increasingly important for you as researchers, regardless of field
- Making research more reproducible benefits all of us
- We have a strong interest in empowering trainees to gain data science skills
- You might find that you really enjoy it

# How did we learn R? {background-color="#2c488f"}



## Kevin's History

- Kid:
    - Computers around
    - Lots of programming
- College:
    - Biology
- Graduate student:
    - Refused to take statistics
    - Did a little "analysis automation" with scripts
    - Tried R a few times and gave up -- "What am I supposed to do with this: `>`?"


## Kevin's History

- Post-doc:
    - SAS / SPSS / Matlab
    - Tried R a few more times and gave up
        - Needed to do something that was only available in R
        - Learned enough R to do that one thing
        - Liked it and learned more
- Faculty member:
    - Didn't want to pay for SAS / SPSS
    - Didn't want to tie students to non-free software


## First script {.smaller}

```{r}
#| eval: false
#| echo: true

#file /Users/kmm/Documents/Work/Projects/Bone/g16_bone - 20 mo (Middleton)/R/qvalue analysis.r

# Created: 31 Jan 2006 by Kevin Middleton
# Modifications:
#		19 Feb 2006: Added expected null line to histogram.
#		20 Feb 2006: Added "prob=TRUE" and "las=1" arguments to
#			hist plotting function. Histogram is now relative (1=null).
#			pi0 now easy to plot on histogram with abline(h=pi0)
#		21 Feb 2006: Added analyses 2 (bootstrapping) & 3 (lambda = 0).
#		22 Mar 2006: Added comparative analyses at end (robust, various FDR)
#		10 Aug 2006: Now loading pvalues_final.txt

# False discovery rate on g16 bone p values
# Data are derived from final tables (1 & 3) of g16 bone ms.

library(qvalue)

setwd(
  "/Users/kmm/Documents/Work/Projects/Bone/g16_bone - 20 mo (Middleton)/R"
)

# Read the data in to a vector
# bonep_reduced_further.txt is a plain text file with 1 p value per row
# 	This file was created using the p values from the final tables of the
#	g16 bone manuscript
pvals <- scan("pvalues_final.txt")

pvals <- sort(pvals, decreasing = TRUE)

# Check distribution of p values.
#pdf(file="/Users/kmm/Documents/Work/R/FDR/hedenfalk/g16_histogram2.pdf", height=6, width=6, horizontal=F)
hist(
  pvals,
  breaks = 20,
  prob = TRUE,
  las = 1,
  xlab = "p value",
  ylab = "Relative Frequency",
  main = ""
)
abline(h = 1, col = 4, lty = 2, lwd = 2)
abline(h = .704, col = "red", lty = 3, lwd = 2)
#dev.off()
#abline(h=length(pvals)/20,col=4,lty=2)

######## ANALYSIS 1 - DEFAULT PARAMETERS ########
# Perform FDR using default parameters:
#	lambda = seq(0, 0.9, 0.05)
#	pi0.method = "smoother"
#	fdr.level = NULL
#	robust = FALSE
#	smooth.df = 3
#	smooth.log.pi0 = FALSE
qobj.def <- qvalue(pvals)

# Open a new plotting window
quartz(width = 7, height = 7)

# Plot the qvalue diagnostic graphs
qplot(qobj.def)

# Print out qvalue summary information
qsummary(qobj.def)

# Need to learn what this value represents
1 - qobj.def$pi0

# Calculate new alpha level
max(qobj.def$pvalues[qobj.def$qvalues <= 0.05])

# Write output to text file
qwrite(qobj.def, filename = "qvalue_default_settings.txt")


######## ANALYSIS 2 - pi0.method = "bootstrap" ########
# Dabney & Storey (qvalue manual) say that for small numbers of p values
#	the "bootstrap" methods described in Storey et al. (2004) is a better
#	estimator of pi0

# Perform FDR with bootstrapping:
#	pi0.method = "bootstrap"
qobj.boot <- qvalue(pvals, pi0.method = "bootstrap")

# Open a new plotting window
quartz(width = 7, height = 7)

# Plot the qvalue diagnostic graphs
#pdf(file="/Users/kmm/Documents/Work/R/FDR/hedenfalk/g16_qplot.pdf", height=6, width=6, horizontal=F)
qplot(qobj.boot)
#dev.off()

# Print out qvalue summary information
qsummary(qobj.boot)

# Need to learn what this value represents
1 - qobj.boot$pi0

# Calculate new alpha level
max(qobj.boot$pvalues[qobj.boot$qvalues <= 0.05])

# Write output to text file
qwrite(qobj.boot, filename = "qvalue_bootstrap.txt")


######## ANALYSIS 3 - Benjamini & Hochberg (1995) ########
# Dabney & Storey (qvalue manual) state that setting lamdba = 0
#	will force pi0 = 1, approximating the assumptions implicit in
# 	Benjamini & Hochberg (1995)

# Perform FDR with lambda = 0:
#	lambda = 0
qobj.bh <- qvalue(pvals, lambda = 0)

# Open a new plotting window
quartz(width = 7, height = 7)

# Plot the qvalue diagnostic graphs
qplot(qobj.bh)

# Print out qvalue summary information
qsummary(qobj.bh)

# Need to learn what this value represents
1 - qobj.bh$pi0

# Calculate new alpha level
max(qobj.bh$pvalues[qobj.bh$qvalues <= 0.05])

# Write output to text file
qwrite(qobj.bh, filename = "qvalue_bh.txt")


qobj.boot <- qvalue(pvals, pi0.method = "bootstrap", fdr.level = 0.05)
qobj.boot.robust <- qvalue(
  pvals,
  pi0.method = "bootstrap",
  robust = TRUE,
  fdr.level = 0.05
)
qobj.lambda0_0.05 <- qvalue(pvals, lambda = 0, fdr.level = 0.05)
qobj.lambda0_0.10 <- qvalue(pvals, lambda = 0, fdr.level = 0.10)
qwrite(qobj.boot, file = "qobj.boot.txt")
qwrite(qobj.boot.robust, file = "qobj.boot.robust.txt")
qwrite(qobj.lambda0_0.05, file = "qobj.lambda0_0.05.txt")
qwrite(qobj.lambda0_0.10, file = "qobj.lambda0_0.10.txt")
```


## When did I feel comfortable teaching with R?

- 2010: First graduate course (CSU San Bernardino)
- 2011-2012: Undergraduate courses (CSU San Bernardino)
- 2016: First graduate course at MU
- 2017: Start co-teaching *Quantitative Methods in Life Sciences* with Libby (almost every year since)
- 2025, 2026: Developing undergraduate courses in Biological Sciences


## Libby's History

- No computer science courses
- First home computers in middle school, first laptop when I started grad school
- Lots of math in college
  - Linear algebra, differential equations, calculus-based statistics


## Libby's History

In Graduate School:

- Interested in data analysis & simulation modeling
- Took graduate-level Multivariate Statistics (SAS)
- Took graduate-level Computer-Intensive Statistical Methods (S-PLUS)


## Simulation Model Dissertation Chapter {.smaller}

```{r}
#| echo: true
#| eval: false

set.seed(19)

#inputs
patch.N <- 7 #number of patches

n <- matrix(100, patch.N) #number of individuals in patch #needs to be matrix

# program to set up initial covariance matrices and generate initial population
# Given.Hmatrix is a dataframe with genetic correlations in upper diagonal
# heritabilities along the diagonal and phenotypic correlations in the lower diagonal

N <- 4 # Number of traits

############create Given.Hmatrix################
Given.Hmatrix <- matrix(0, N, N)
# a=liability, b=gxe liability, p=prop allocate, z=gxe allocate

h2 <- c(0.51, 0.52, 0.53, 0.54) # heritabilities of traits a, b, p, z
Ra.ab <- 0.1 #liability and gxe liability
Rp.ab <- 0.12
Ra.ap <- 0.2 #liability and prop allocate
Rp.ap <- 0.22
Ra.az <- 0.3 #liability and gxe allocate
Rp.az <- 0.33
Ra.bp <- 0.4 #gxe liability and prop allocate
Rp.bp <- 0.44
Ra.bz <- 0.5 #gxe liability and gxe allocate
Rp.bz <- 0.55
Ra.pz <- 0.6 #prop allocate and gxe allocate
Rp.pz <- 0.66

diag(Given.Hmatrix) <- h2 #puts above into Given.Hmatrix
Given.Hmatrix[1, 2] <- Ra.ab
Given.Hmatrix[1, 3] <- Ra.ap
Given.Hmatrix[1, 4] <- Ra.az
Given.Hmatrix[2, 3] <- Ra.bp
Given.Hmatrix[2, 4] <- Ra.bz
Given.Hmatrix[3, 4] <- Ra.pz

Given.Hmatrix[2, 1] <- Rp.ab
Given.Hmatrix[3, 1] <- Rp.ap
Given.Hmatrix[4, 1] <- Rp.az
Given.Hmatrix[3, 2] <- Rp.bp
Given.Hmatrix[4, 2] <- Rp.bz
Given.Hmatrix[4, 3] <- Rp.pz

Given.Hmatrix <- data.frame(Given.Hmatrix)
H2 <- Given.Hmatrix # Given.Hmatrix CW W5 ICG GR


# Initial Additive genetic means for each patch
Mean.A <- matrix(0, patch.N, N) # Additive genetic means
Mean.E <- matrix(0, patch.N, N) # Environmental mean

# Phenotypic Covariance matrix
# Note that initial covariances set to 1 (arbitrary)
CovP <- matrix(1, N, N) # Phenotypic variances
diag(CovP) <- c(1.8, 1.53 / 16, 0.67, 0.61) # Sets diagonal elements = variances

# Establish CovA from h2 and CovP and CovE from CovA and CovP
CovA <- matrix(0, N, N)
CovE <- matrix(0, N, N)

for (i in 1:N) {
  CovA[i, i] <- CovP[i, i] * H2[i, i] # Vp*h2
  CovE[i, i] <- CovP[i, i] - CovA[i, i] # Variances See  p77 of Roff 1997
  if (CovE[i, i] < 0) stop(print(c("CovE cannot be", i, j, CovE[i, i])))
}

# Phenotypic and genetic covariances and correlations
N.minus.1 <- N - 1
for (i in 1:N.minus.1) {
  jj <- i + 1
  for (j in jj:N) {
    CovP[i, j] <- H2[j, i] * sqrt(CovP[i, i] * CovP[j, j]) # Phenotypic covariance p76 Roff 1997
    CovP[j, i] <- CovP[i, j]
    CovA[i, j] <- H2[i, j] * sqrt(CovA[i, i] * CovA[j, j]) # Ra*sqrt(VaxVay), p76
    CovA[j, i] <- CovA[i, j]
    CovE[i, j] <- CovP[i, j] - CovA[i, j] #  p76
    CovE[j, i] <- CovE[i, j]
  }
}

################### SLIBBY PROGRAM ###################

for (i in 1:patch.N) {
  #Set Resource level for patches
  muR <- 0 #mean resource level
  sdR <- 1 #sd for resource level
  X <- rnorm(patch.N, muR, sdR) #generates resource levels for patches
  X <- as.matrix(X)
  # Generate additive genetic and environmental values using bivariate
  Trait.E <- rmvnorm(n[i, ], mean = Mean.E[i, ], cov = CovE, d = N) #environmental values
  Trait.A <- rmvnorm(n[i, ], mean = Mean.A[i, ], cov = CovA, d = N) #additive genetic values

  # Phenotypic value = sum of above
  Trait.P <- Trait.E + Trait.A #phenotypic values

  # Wing morph

  Prop.LW <- 0.83 # Proportion LW is proportion greater than threshold
  Z.LW <- -qnorm(Prop.LW) #need to specify mean?

  a <- Trait.P[, 1]
  b <- Trait.P[, 2]
  L <- a + b * X[i, ] #liability
  Wing <- L
  Wing[L < Z.LW] <- 0 #sw
  Wing[L > Z.LW] <- 1 #lw

  #*************#RL shift###not yet coded

  # Proportion Allocated
  p <- Trait.P[, 3]
  z <- Trait.P[, 4]
  PA <- p + z * X[i, ] #p(allocate)

  PA[PA < 0] <- 0 # Set p = 0 if < 0
  PA[PA > 1] <- 1 # Set p = 1 if < 1

  Traits.Allocate <- cbind(Wing, PA)

  # Set all SW as having a PA of 0
  Traits.Allocate[Traits.Allocate[, 1] < 0.5, 2] <- 0

  ####mate individuals randomly within patches###
  pop.means <- as.matrix(colMeans(Trait.A))

  bv.a <- (Trait.A[, 1] + pop.means[1, 1]) / 2
  bv.b <- (Trait.A[, 2] + pop.means[2, 1]) / 2
  bv.p <- (Trait.A[, 3] + pop.means[3, 1]) / 2
  bv.z <- (Trait.A[, 4] + pop.means[4, 1]) / 2
  BV <- cbind(Traits.Allocate, bv.a, bv.b, bv.p, bv.z) #have bv migrate? or traits all?

  #separate LW and SW
  BV.SW <- BV[BV[, 1] == 0, ]
  BV.LW <- BV[BV[, 1] == 1, ]
  n.lw <- length(BV.LW[, 1]) #number of lw in patch
  n.sw <- n[i, ] - n.lw
  ######MIGRATION########################################################################

  #assign 0 prob to all SW
  #m.vector			<-rep(0,length(BV.SW[,1]))
  #BV.SW				<-cbind(BV.SW,m.vector)

  #determine migrants
  #pm					<-0.5 #begin with constant proportion- to be function of PA[prob in sample]
  #m.vector			<-c(rep(1,pm*n.lw),rep(0,(1-pm)*n.lw))
  #m.vector			<-sample(m.vector)
  #BV.LW				<-cbind(BV.LW,m.vector)

  #separate migrants and nonmigrants
  #BV.nmig						<-BV.LW[BV.LW[,7]==0,]
  #BV.mig							<-BV.LW[BV.LW[,7]==1,]

  #####combine sw with non migrants##
  #BV.Total<-rbind(BV.SW,BV.nmig)

  #####send migrants to migrant pool# not yet coded
  #Mig.Pool<-rbind(

  ####redistribute migrants## not yet coded
  ###############################END MIGRATION#########################3333

  ##reproduce##based on PA
  temp <- as.matrix(colMeans(BV.LW))
  Mean.bv.lw <- matrix(0, 1, N)
  Mean.bv.lw <- (c(temp[3, ], temp[4, ], temp[5, ], temp[6, ]))

  temp <- as.matrix(colMeans(BV.SW))
  Mean.bv.sw <- matrix(0, 1, N)
  Mean.bv.sw <- (c(temp[3, ], temp[4, ], temp[5, ], temp[6, ]))

  growth.sw <- 1.5
  growth.lw <- 1.1

  n[i, ] <- (growth.sw * n.sw) + (growth.lw * n.lw)
  Mean.A[i, ] <- (((growth.sw * n.sw) * Mean.bv.sw) +
    ((growth.lw * n.lw) * Mean.bv.lw)) /
    n[i, ]
}

```


## Libby's History

- Used R to analyze all dissertation data
- Got a computational genomics post-doc
    - Learned Perl, bash, remote computers, etc.
    - Used R every day


## Our Shared History

- Both learning R at the same time
- Helped each other
- Worked together
- Sent help resources to each other
- Developed courses together


## Roadmap to Feeling Confident

1. Practice (Regularly)

> What is something you are good at?


# Some questions about you {background-color="#2c488f"}


## Outline for today

- Example use cases
- Introduction to R/RStudio & Reproducibility
- Practice
- Roadmaps for different levels of learners
- Roadmaps for different specific skills


## Things you can do in R/RStudio

- Making blank data sheets
- Data wrangling (joins, pivots, finding and correcting errors)
- Statistics and data analysis
- Publication-ready data visualizations
- Reproducible manuscripts, reports, books, dissertations
- Slides
- CVs and Websites


## What tools do we use?

- [R](https://www.r-project.org/): Statistical programming language
- [RStudio](https://posit.co/download/rstudio-desktop): Graphical user interface for R
- [Quarto](https://quarto.org/){target="_blank"}: Markdown language for mixing code, text, and visualization (included with RStudio)


## Introduction to R/RStudio & Reproducibility

- R is a general purpose programming language (that is optimized for statistics and data visualization)
- You can do all of these things in other languages (some easier and many more difficult)
- R enforces an "ask for what you want" statistical approach

> The term "environment" is intended to characterize it as a fully planned and coherent system, rather than an incremental accretion of very specific and inflexible tools, as is frequently the case with other data analysis software.


## Modern scientists are programmers

- Most code written for yourself or your collaborators
- 95% of your analysis time is spent _not_ doing statistics.
    - Wrangling data
    - Plotting data
- Invest time now in learning the logic of programming. Save time later.
- Reproducibility. You always know what you did.
    - Code with comments reads as a step-by-step of your analysis
- This is a process.
    - It takes time to become comfortable.


## What is reproducible research?

> "By reproducible, I mean that your work can be repeated by other researchers and they can arrive at the same results. For this to be the case, your work must be well documented, and your methods, code, and data all need to be available so that other researchers have the materials to reproduce everything."

[Buffalo (2015)]{style="float: right;"}


## R Markdown (`.rmd`) / Quarto (`.qmd`)

- Text-based markup languages (i.e., HTML)
- Render to different formats (PDF, HTML, Word)
- Inline code
    - Your text and analysis coexist together in one file
- Resources
    - Markdown information
    - Guides and Tutorials


## Markdown-based analyses

- More than just comments
- Add context to your analysis
- Narrative, figures, tables, etc. all in one document



# Let's practice {background-color="#2c488f"}


## These materials are online

- Slides: [https://tinyurl.com/GPC-R-Intro](https://tinyurl.com/GPC-R-Intro)
- Data: [https://tinyurl.com/NHANES-Data](https://tinyurl.com/NHANES-Data)


## Discussion

- What do you need for what you want to use R for?


# Roadmaps for different levels of learners {background-color="#2c488f"}

## Resources

[https://tinyurl.com/QMLS-Resources](https://tinyurl.com/QMLS-Resources)


## Roadmaps for different levels of learners

QR code for each then slide with grid of instructions

# Roadmaps for different specific skills {background-color="#2c488f"}
